---
title: "Time Series with HLS"
author: "Michele Tobias, PhD"
format: html
editor: visual
---

## Time Series with HLS & STAC Catalogs

This notebook will explore doing a time series analysis with HLS data for 10 Mile Dunes in northern California.

### Set Up

#### Libraries

First, we need to install any R libraries we want to use. You only need to do this once per computer.

```{r}
# Install Packages - only need to do this once 
install.packages(c("rstac", "terra", "RColorBrewer", "getPass", "earthdatalogin"))
```

Next, we need to load those libraries. You might assume that if we install the library, we want to use it, but R doesn't automatically load libraries you just installed.

```{r}
# Load Libraries 
library(rstac)          # interact with stac catalogs 
library(terra)          # work with raster data 
library(RColorBrewer)   # color pallets from Color Brewer, a library of colors designed for optimized readability in cartograhy 
library(getPass)        # allows us to give the code our password without it printing to the console 
library(earthdatalogin) # lets us authenticate to access NASA data
```

#### Authenticate and connect to the NASA Earth Data STAC Catalog

Store your user name and password in the R session but not in your code:

```{r}
earthdata_username <- getPass("Earthdata User:")
earthdata_password <- getPass("Earthdata Password:")
```

Now we run the authentication function inputting our username and password:

```{r}
earthdatalogin::edl_netrc(
  username = earthdata_username, 
  password = earthdata_password)
```

## Search HLS

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
